---
title: "Sample analysis: Streptomycin on TB"
author: Matt Broerman
format: 
  pdf:
    toc: true
  # html:
  #   embed-resources: true
  #   code-fold: true
  #   toc: true
---

<!-- This is a comment.-->
<!-- Note that I am outputting two formats from one "source" -->
<!-- Each format has it's own settings, but some are only relevant to one -->

## Set Up

```{r message=FALSE}

library(medicaldata)

library(dplyr)
library(tidyr)
library(janitor)

library(knitr)
library(gt)
library(gtsummary)

strep_tb <- medicaldata::strep_tb |> as_tibble()
```

<!-- Note that I rearranged the content from demo_2_sum_sol.R. -->
<!-- Note that I used message=FALSE to keep messages out of the compiled report -->

## Baseline vs outcome

In reviewing this study, notice that outcomes are both improvements, but different effect sizes. 

```{r}
strep_base <-
  strep_tb |>
  select(arm, baseline_condition, gender, rad_num) |>
  separate(baseline_condition, into = c("baseline_num", "baseline_cond")) |>
  mutate(baseline_num = as.numeric(baseline_num))

strep_base_sum <-
  strep_base |>
  group_by(arm) |>
  summarise(
    n_pt = n(),
    mean_baseline = mean(baseline_num),
    mean_outcome = mean(rad_num)
  )

strep_base_sum |>
  kable(digits = 3)
```

## Treatment vs outcome

The point above is important for interpreting the tables below.

<!-- Note the first table is not a "kable" -->

```{r}
strep_tb |> 
  count(radiologic_6m, improved)

strep_tb |>
  tabyl(arm, improved) |> 
  kable(caption = "Improved F/T")
```

That is, strictly speaking, on average both treatment groups "improved", but for the category "improved" defined on a likert scale, more of the treated vs control were "improved". 
