---
title: "Demo: Microbiome Analysis with Phyloseq"
author: "Matt Broerman"
format: 
  html:
    embed-resources: true
---

## Introduction

In this demonstration, you will work with microbiome data using the `phyloseq` package in R. You will explore the data, manipulate it, and create visualizations to gain insights into microbial communities.

## Setup

Load the necessary libraries and data.

```{r message=FALSE}
# Load necessary packages
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("phyloseq", quietly = TRUE)) {
     BiocManager::install("phyloseq")
}

library(phyloseq)
library(tidyverse)
library(here)

# Load the phyloseq object
ps <- readRDS(here("data", "lipuma_cf.rds"))
```

## Questions and Exercises

### Question 1

Understanding the phyloseq object

a) What are the components of the phyloseq object ps? List them.


The `phyloseq` object `ps` consists of:

- OTU table (`otu_table`)
- Sample data (`sample_data`)
- Taxonomy table (`tax_table`)



b) How many samples and taxa are in the dataset?


```{r}
nsamples(ps)
ntaxa(ps)
```



### Question 2

Exploring Sample Metadata

a) View the first few rows of the sample data. What variables are included in the sample metadata?


```{r}
head(sample_data(ps))
```
Variables include sample ID, subject ID, sample type, treatment group, etc.



b) How many samples are sputum and how many are saliva?


```{r}
table(sample_data(ps)$sample_type)
```



### Question 3

Subsetting the Data

a) Create a subset of the data that includes only sputum samples. Assign this subset to a new phyloseq object called ps_sputum.



```{r}
ps_sputum <- subset_samples(ps, sample_type == "Sputum")
```




b) How many samples are in ps_sputum?


```{r}
nsamples(ps_sputum)
```



### Question 4

Taxonomic Summaries

a) Generate a bar plot showing the relative abundance of phyla across all samples.


```{r}
ps |>
  transform_sample_counts(function(x) x / sum(x)) |>
  plot_bar(fill = "phylum")
```



b) Which phylum is most abundant in the dataset?


By examining the bar plot, you can identify the most abundant phylum.



### Question 5

Alpha Diversity Analysis

a) Calculate the Shannon diversity index for each sample and add it to the sample data.


```{r}
shannon_values <- estimate_richness(ps, measures = "Shannon")
sample_data(ps)$Shannon <- shannon_values$Shannon
```



b) Plot the Shannon diversity index by sample type (sputum vs. saliva). What can you conclude from the plot?


```{r}
ggplot(sample_data(ps), aes(x = sample_type, y = Shannon)) +
  geom_boxplot() +
  labs(title = "Shannon Diversity by Sample Type")
```
You might conclude that one sample type has higher diversity than the other.



### Question 6

Beta Diversity and Ordination

a) Perform a Principal Coordinates Analysis (PCoA) using Bray-Curtis distance on the samples.


```{r}
ordination <- ordinate(ps, method = "PCoA", distance = "bray")
```



b) Create an ordination plot colored by sample type.


```{r}
plot_ordination(ps, ordination, color = "sample_type") +
  geom_point(size = 3) +
  labs(title = "PCoA of Samples")
```



c) Do sputum and saliva samples cluster separately?


Based on the ordination plot, you can observe whether the samples cluster by type.



### Question 7

Differential Abundance

a) Identify taxa that are significantly different in abundance between sputum and saliva samples at the genus level.


You can use DESeq2 for differential abundance analysis.

```{r message=FALSE}
# Install DESeq2 if not already installed
if (!requireNamespace("DESeq2", quietly = TRUE)) {
    BiocManager::install("DESeq2")
}
library(DESeq2)

# Agglomerate taxa at the genus level
ps_genus <- tax_glom(ps, taxrank = "genus")

# Convert to DESeq2 object
dds <- phyloseq_to_deseq2(ps_genus, ~ sample_type)

# Run DESeq2 analysis
dds <- DESeq(dds)

# Get results
res <- results(dds)

# Subset significant results
res_sig <- res[which(res$padj < 0.05), ]

# View significant taxa
res_sig
```



b) Which genera are more abundant in sputum compared to saliva?


Genera with positive log2 fold change values are more abundant in sputum.



### Question 8

Custom Plotting

a) Create a heatmap of the top 20 most abundant genera across samples.


```{r}
# Transform counts to relative abundances
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

# Select top 20 genera
top20_taxa <- names(sort(taxa_sums(ps_rel), decreasing = TRUE))[1:20]
ps_top20 <- prune_taxa(top20_taxa, ps_rel)

# Plot heatmap
plot_heatmap(ps_top20, sample.label = "samp_cod", taxa.label = "genus")
```



b) Interpret the heatmap. Are there any patterns you can observe?


Interpretation will depend on the observed patterns, such as clustering of samples or taxa.


